---
title: "Deprivation"
output:
  html_document:
    df_print: paged
---
# Activity by Board of Treatment and Deprivation

SIMD 
https://www.opendata.nhs.scot/dataset/scottish-index-of-multiple-deprivation

ScotGOv
https://www.gov.scot/collections/scottish-index-of-multiple-deprivation-2020/

"The Scottish Index of Multiple Deprivation is a relative measure of deprivation 
across 6,976 small areas (called data zones). If an area is identified as 
‘deprived’, this can relate to people having a low income but it can also mean 
fewer resources or opportunities. SIMD looks at the extent to which an area is deprived across seven domains: income, employment, education, health, access to services, crime and housing."

```{r warning=FALSE}
library(tidyverse)
```

```{r message=FALSE, warning=FALSE}
deprivation <- read_csv("../clean_data/deprivation_without_duplicates.csv")
```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
head(deprivation)
glimpse(deprivation)
str(deprivation)
skimr::skim(deprivation)
```

## AdmissionType, Quarter
Line graph confirms similar trends across all departments.
Apparent lower number of stays post 2020.
```{r}
admission_quarter <- deprivation %>% 
  select(admission_type, year_quarter, stays) %>% 
  group_by(admission_type, year_quarter) %>% 
  summarise(total_stays = sum(stays)) %>% 
  ungroup()

admission_quarter %>% 
  ggplot(aes(x = admission_type, y = total_stays, fill = admission_type)) +
  geom_col()

admission_quarter %>% 
  ggplot(aes(x = year_quarter, y = total_stays,
             colour = admission_type, group = admission_type)) +
  geom_line() +
  geom_point()
```

## SIMD, Quarter
SIMD ranks data zones into quintiles, from most deprived (ranked 1) to least
deprived (ranked 5).
Line graph confirms similar trends across all SIMD quintiles.
Apparent lower number of stays post 2020 for all groups.

```{r}
deprivation %>% 
  select(simd, year_quarter, stays) %>%
  group_by(simd, year_quarter) %>% 
  summarise(total_stays = sum(stays)) %>%
  ungroup() %>% 
  ggplot(aes(x = year_quarter, y = total_stays,
             colour = factor(simd), group = simd)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, NA),
                     labels = scales::comma) +
  scale_colour_discrete(name = "SIMD") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Stays by Scottish Index of Multiple Deprivation (SIMD)",
    x = "\nQuarter ending",
    y = "Total Number of Stays in Days\n")


deprivation %>% 
  select(simd, year_quarter, average_length_of_stay) %>%
  group_by(simd, year_quarter) %>% 
  summarise(mean_stays = mean(average_length_of_stay)) %>%
  ungroup() %>% 
  ggplot(aes(x = year_quarter, y = mean_stays,
             colour = factor(simd), group = simd)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, NA),
                     labels = scales::comma) +
  scale_colour_discrete(name = "SIMD") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Average Length of Stays by SIMD",
    x = "\nQuarter ending",
    y = "Total Number of Stays in Days\n")
```

Length of stay
```{r}
# Mean length of stay
# No clear trend.
# Range is around 1 day
deprivation %>% 
  select(year_quarter, simd, average_length_of_stay) %>% 
  group_by(year_quarter) %>% 
  summarise(mean_stay = mean(average_length_of_stay, na.rm = TRUE)) %>% 
  ungroup() %>%
  ggplot(aes(x = year_quarter, y = mean_stay)) +
  geom_col(fill = "seagreen") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Mean Length of Stay",
    x = "",
    y = "Days")
```

