---
title: "Analysis: Sex"
output:
  html_document:
    df_print: paged
---

# Hospital Activity and Patient Demographics

https://www.opendata.nhs.scot/dataset/inpatient-and-daycase-activity/resource/00c00ecc-b533-426e-a433-42d79bdea5d4

# Activity by Board of Treatment, Age and Sex
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(infer)
```

```{r message=FALSE, warning=FALSE}
age_sex <- read_csv("../clean_data/age_and_sex_without_duplicates.csv")
```

```{r eval=FALSE, include=FALSE}
glimpse(age_sex)

str(age_sex)

skimr::skim(age_sex)
```

# Sex
```{r message=FALSE}
# Number of "stays" (total number of days patients spend in hospital during a continuous inpatient stay)
# Consistently Female > Male by around 50,000 per quarter
age_sex %>%
  select(year_quarter, sex, stays) %>%
  group_by(year_quarter, sex) %>%
  summarise(stays_count = sum(stays)) %>%
  ungroup() %>%
  ggplot(aes(x = year_quarter, y = stays_count, colour = sex, group = sex)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, NA),
                     labels = scales::label_number(
                       scale_cut = scales::cut_short_scale())) +
  theme_minimal(default_font_base_size) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  labs(title = "Number of Stays in Hospital",
       x = "",
       y = "Total Number of Stays\n",
       colour = "")

# Mean length of stay
# No clear trend...maybe Female > Male on average
# Range is small
age_sex %>% 
  select(year_quarter, sex, average_length_of_stay) %>% 
  group_by(year_quarter, sex) %>% 
  summarise(mean_stay = mean(average_length_of_stay, na.rm = TRUE)) %>% 
  ungroup() %>%
  ggplot(aes(x = year_quarter, y = mean_stay,
             group = sex, colour = sex)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, NA)) +
  theme_minimal(base_size = default_font_base_size) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  labs(title = "Length of Stay",
       x = "",
       y = "Average Number of Days\n",
       colour = "")
```

# Question 1: To what extent are the ‘winter crises’ reported by the media real?

```{r}
age_sex %>% 
  group_by(winter) %>% 
  summarise(mean_stays = mean(stays, na.rm = TRUE),
            mean_length_of_stay = mean(average_length_of_stay, na.rm = TRUE))
```
Doesn't look like a significant difference.

### HYPOTHESIS TESTING

h0: the number of stays (both sexes total) is not significantly greater in winter quarters relative to non-winter quarters
h1: the number of stays is greater in winter quarters

```{r}
# Boxplot
age_sex %>% 
  ggplot(aes(x = stays, y = winter, fill = winter)) +
  geom_boxplot()

# Set up null distriburtion
null_dist_winter <- age_sex %>% 
  specify(stays ~ winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

# Calculate the observed stat
obs_stat_winter <- age_sex %>% 
  specify(stays ~ winter) %>% 
  calculate(stat = "diff in means", order = c(TRUE, FALSE))
            
# Visualise confidence interval 
null_dist_winter %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_winter, direction = "right")

# Find p-value
pvalue_winter <- null_dist_winter %>% 
  get_p_value(obs_stat = obs_stat_winter, direction = "right")
pvalue_winter

```
With a p-value around 0.25, we fail to reject the null hypothesis.
The number of hospital stays is, on average, greater in winter, but not significantly so.


# Question 2: How has the Covid-19 pandemic affected provision of acute care in Scotland?
- Demographic: Who is most affected by this issue? (Who should be targeted with efforts?)

## Overall
```{r}
age_sex %>% 
  group_by(post_covid) %>% 
  summarise(mean_stays = mean(stays, na.rm = TRUE),
            mean_length_of_stay = mean(average_length_of_stay, na.rm = TRUE))
```
Drop in the number of stays....looks significant.
Increase in length of stay....significant? Doubtful.

*Hypthesis Test - mean stays*
h0: There is no significant decrease post-covid in the overall number of stays in hospital.
h1: There are significantly fewer stays in hospital post-covid.
```{r}
age_sex %>% 
  ggplot(aes(x = stays, y = post_covid, fill = post_covid)) +
  geom_boxplot()

null_distribution <- age_sex %>%
  specify(response = stays, explanatory = post_covid) %>% 
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

observed_stat <- age_sex %>% 
  specify(stays ~ post_covid) %>% 
  calculate(stat= "diff in means", order = c(TRUE, FALSE))

observed_stat

null_distribution %>% 
  visualise(bins = 30) +
  shade_p_value(obs_stat = observed_stat, direction = "left")

null_distribution %>% 
  get_p_value(obs_stat = observed_stat, direction = "left")
```
*Very small p-value leads us to reject the null hypothesis.*
*There are significantly fewer stays in hospital post-covid.*  Why?

*Hypthesis Test - mean length of stays*
h0: There is no significant increase post-covid in the overall length of stays in hospital.
h1: Stays in hospital are significantly longer post-covid.
```{r}
age_sex %>% 
  ggplot(aes(x = average_length_of_stay, y = post_covid, fill = post_covid)) +
  geom_boxplot()

null_distribution <- age_sex %>%
  specify(response = average_length_of_stay, explanatory = post_covid) %>% 
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

observed_stat <- age_sex %>% 
  specify(average_length_of_stay ~ post_covid) %>% 
  calculate(stat= "diff in means", order = c(TRUE, FALSE))

observed_stat

null_distribution %>% 
  visualise(bins = 30) +
  shade_p_value(obs_stat = observed_stat, direction = "right")

null_distribution %>% 
  get_p_value(obs_stat = observed_stat, direction = "right")
```
*Small p-value (approx 0.005) leads us to reject the null hypothesis.*
*Stays in hospital are significantly longer post-covid.*
*It should be noted that this statistical significance translates as around 3 hours extra, on average*




## Post-covid changes of different sexes
```{r}
age_sex %>% 
  group_by(sex, post_covid) %>% 
  summarise(mean_stays = mean(stays, na.rm = TRUE),
            mean_length_of_stay = mean(average_length_of_stay, na.rm = TRUE))
```
Post-covid, the number of stays in both groups has decreased by a lot. Has one sex been affected more than the other?
Post-covid, the length of stays in both groups has increased by a small amount.  Has one sex been affected more than the other?

```{r message=FALSE, warning=FALSE}
# Wrangling data set
male_stays <- age_sex %>% 
  filter(sex == "Male") %>% 
  rename(male_stays = stays,
         average_length_of_male_stay = average_length_of_stay) %>% 
  group_by(year_quarter, hb, admission_type, age, winter, post_covid) %>% 
  summarise(male_stays = sum(male_stays),
            average_length_of_male_stay = sum(average_length_of_male_stay)) %>% 
  ungroup() %>% 
  mutate(id = str_c(year_quarter, hb, admission_type, age))

female_stays <- age_sex %>%
  filter(sex == "Female") %>%
  rename(female_stays = stays,
         average_length_of_female_stay = average_length_of_stay) %>%
  group_by(year_quarter, hb, admission_type, age, winter, post_covid) %>% 
  summarise(female_stays = sum(female_stays),
            average_length_of_female_stay = sum(average_length_of_female_stay)) %>% 
  ungroup() %>% 
  mutate(id = str_c(year_quarter, hb, admission_type, age))

age_sex_stay_deltas <- full_join(male_stays, female_stays, by = "id") %>%
  mutate(
    female_stays = case_when(female_stays > 0 ~ female_stays,
      .default = 0
    ),
    average_length_of_female_stay =
      case_when(average_length_of_female_stay > 0 ~ average_length_of_female_stay,
        .default = 0
      ),
    male_stays = case_when(male_stays > 0 ~ male_stays,
      .default = 0
    ),
    average_length_of_male_stay =
      case_when(average_length_of_male_stay > 0 ~ average_length_of_male_stay,
        .default = 0
      )
  ) %>%
  rename(
    winter = winter.x,
    post_covid = post_covid.x
  ) %>%
  mutate(winter = case_when(winter == TRUE ~ TRUE,
    winter == FALSE ~ FALSE,
    .default = winter.y
  )) %>%
  mutate(post_covid = case_when(post_covid == TRUE ~ TRUE,
    post_covid == FALSE ~ FALSE,
    .default = post_covid.y
  )) %>%
  mutate(
    more_female_stays = female_stays - male_stays,
    females_stay_longer = average_length_of_female_stay - average_length_of_male_stay
  ) %>%
  select(more_female_stays, females_stay_longer, winter, post_covid)
```


### Number of hospital stays by sex pre/post covid
Hypothesis test:
h0: There is no significant change post-covid in the difference between the number of male and female stays in hospital.
h1: There is a significant change post-covid in the difference between the number of male and female stays in hospital.
```{r}
age_sex_stay_deltas %>% 
  ggplot(aes(x = more_female_stays, y = post_covid, fill = post_covid)) +
  geom_boxplot()

null_distribution <- age_sex_stay_deltas %>%
  specify(response = more_female_stays, explanatory = post_covid) %>% 
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

observed_stat <- age_sex_stay_deltas %>% 
  specify(more_female_stays ~ post_covid) %>% 
  calculate(stat= "diff in means", order = c(TRUE, FALSE))

observed_stat

null_distribution %>% 
  visualise(bins = 30) +
  shade_p_value(obs_stat = observed_stat, direction = "both")

null_distribution %>% 
  get_p_value(obs_stat = observed_stat, direction = "both")
```
*p in region of 0.3 so fail to reject the null hypothesis*
*There is no significant change post-covid in the difference between the number of male and female stays in hospital.*

### Length of hospital stays by sex pre/post covid
Hypothesis test 2:
h0: There is no significant change post-covid in the difference between the length of stays of men and women staying in hospital.
h1: There is a significant change post-covid in the difference between the length of stays of men and women staying in hospital.

```{r}
age_sex_stay_deltas %>% 
  ggplot(aes(x = females_stay_longer, y = post_covid, fill = post_covid)) +
  geom_boxplot()

null_distribution <- age_sex_stay_deltas %>%
  specify(response = females_stay_longer, explanatory = post_covid) %>% 
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

observed_stat <- age_sex_stay_deltas %>% 
  specify(females_stay_longer ~ post_covid) %>% 
  calculate(stat= "diff in means", order = c(TRUE, FALSE))

observed_stat

null_distribution %>% 
  visualise(bins = 30) +
  shade_p_value(obs_stat = observed_stat, direction = "both")

null_distribution %>% 
  get_p_value(obs_stat = observed_stat, direction = "both")
```
*p in the region of 0.8 so fail to reject the null hyposthesis*
*There is no significant change post-covid in the difference between the length of stays of men and women.*