---
title: "A and E Waiting times"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(lubridate)
library(infer)
waiting_times <- janitor::clean_names(read_csv("../raw_data/monthly_ae_activity_202307.csv"))
names(waiting_times)

# Extract time info
waiting_times_clean <- waiting_times %>% 
  mutate(
    date = ymd(month, truncated = 2)
  ) 
  
# remove unwanted columns
waiting_times_clean <- waiting_times_clean %>%   
  select(-number_within4hours_episode_qf, -number_of_attendances_episode_qf,
         -number_over4hours_episode_qf, -percentage_over8hours_episode_qf,
        -number_over8hours_episode_qf,percentage_over12hours_episode_qf)

waiting_times_clean <- waiting_times_clean 
  
# rename healthboards
waiting_times_clean <- waiting_times_clean %>% 
  mutate(health_board = case_when(
    hbt %in% "S08000015" ~ "Ayrshire and Arran",
    hbt %in% "S08000016" ~ "Borders",
    hbt %in% "S08000017" ~ "Dumfries and Galloway",
    hbt %in% "S08000029" ~ "Fife",
    hbt %in% "S08000019" ~ "Forth Valley",
    hbt %in% "S08000020" ~ "Grampian",
    hbt %in% "S08000031" ~ "Greater Glasgow and Clyde",
    hbt %in% "S08000022" ~ "Highland",
    hbt %in% "S08000032" ~ "Lanarkshire",
    hbt %in% "S08000024" ~ "Lothian",
    hbt %in% "S08000025" ~ "Orkney",
    hbt %in% "S08000026" ~ "Shetland",
    hbt %in% "S08000030" ~ "Tayside",
    hbt %in% "S08000028" ~ "Western Isles",
    T ~ country
  )
  )

# make new column with yearly quarter data
waiting_times_clean <- waiting_times_clean %>% 
  mutate(
    year_quarter = paste0(year(date),
                   " Q",
                   quarter(date)),
    year = year(date),
    quarter = quarter(date)
  )

# group into winter/non-winter based on quarter
waiting_times_clean <- waiting_times_clean %>% 
 mutate(winter = case_when(str_detect(year_quarter, "Q1") ~ "Winter",
                            str_detect(year_quarter, "Q4") ~ "Winter",
                            .default = "Non-Winter"),
         post_covid = year > 2019)

# filter to remove unwanted years
waiting_times_clean <- waiting_times_clean %>% 
  filter(year > 2017)

# make new col with percentage of waits
waiting_times_clean <- waiting_times_clean %>% 
  group_by(year_quarter) %>% 
  mutate(q_percent_within_4_hours = (number_within4hours_all/number_of_attendances_all)* 100,
         q_percent_over_4_hours = (number_over4hours_all/number_of_attendances_all)* 100,
         q_percent_above_8_hours = (number_over8hours_episode/number_of_attendances_all)* 100,
         q_percent_above_12_hours = (number_over12hours_episode/number_of_attendances_all)* 100)

# rename variables to match elsewhere
waiting_times_clean <- waiting_times_clean %>% 
  mutate(
    hb_name = health_board,
    hb = hbt)

# rename and save
waiting_times_condense <- waiting_times_clean

write.csv(waiting_times_condense, file = "../clean_data/waiting_times_condense.csv")


```



# Plot showing percentage of waits within 4 hours
```{r}
waiting_times_clean %>% 
  ggplot(aes(x = year_quarter, y = number_within4hours_episode, fill = winter)) +
  geom_bar(stat = "identity") +
  labs(title = "Seen within 4 hours",
       x = "",
       y = "Episodes\n",
       fill = "") +
  theme_minimal()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_y_continuous(labels = scales::number_format())
```

# Waits over 4 hours

```{r}
waiting_times_condense %>% 
  ggplot(aes(x = year_quarter, y =  number_over4hours_episode, fill = winter)) +
  geom_bar(stat = "identity") +
  labs(title = "Wait > 4 hours",
       x = "",
       y = "Episodes\n",
       fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::number_format())
  
```
# Over 8 hours

```{r}
waiting_times_condense %>% 
  ggplot(aes(x = year_quarter, y =  number_over8hours_episode, fill = winter)) +
  geom_bar(stat = "identity") +
  labs(title = "Wait > 8 hours",
       x = "",
       y = "Episodes\n",
       fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::number_format())
  
```

# Over 12 hours

```{r}

waiting_times_condense <- read_csv("../clean_data/waiting_times_condense.csv")


waiting_times_condense %>% 
  ggplot(aes(x = year_quarter, y = number_over12hours_episode, fill = winter)) +
  geom_bar(stat = "identity") +
  labs(title = "Wait > 12 hours",
       x = "",
       y = "Episodes\n",
       fill = "") +
  theme_minimal()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::number_format())

```

# H testing winter crises

## Wait within 4 hours

h0: the proportion of waiting times that were less than 4 hours was not significantly greater less in winter quarters relative to non-winter quarters
h1: the proportion of waiting times that were less than 4 hours was greater less in winter quarters

```{r}
waiting_times_condense %>% 
  ggplot(aes(y = number_within4hours_episode, x = winter, fill = winter)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
library(infer)

null_dist_within_4hrs <- waiting_times_condense %>% 
  specify(number_within4hours_episode ~ winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
```

```{r}
obs_stat_within_4 <- waiting_times_condense %>% 
  specify(number_within4hours_episode ~ winter) %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
            
obs_stat_within_4
```

```{r}
null_dist_within_4hrs %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_within_4, direction = "left")
  
```

```{r}
pvalue_within_4hr <- null_dist_within_4hrs %>% 
  get_p_value(obs_stat = obs_stat_within_4, direction = "left")


pvalue_within_4hr
```


## Wait over 4 hours

h0: waits over 4 hours were not significantly greater in winter quarters relative to non-winter quarters
h1: the proportion of waits over 4 hours was greater in winter quarters

```{r}

waiting_times_condense %>% 
  ggplot(aes(y = number_over4hours_episode, x = winter, fill = winter)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
null_dist_over_4hrs <- waiting_times_condense %>% 
  specify(number_over4hours_episode ~ winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
```

```{r}
obs_stat_over_4 <- waiting_times_condense %>% 
  specify(number_over4hours_episode ~ winter) %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
            
obs_stat_over_4
```

```{r}
null_dist_over_4hrs %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_over_4, direction = "both")
  
```

```{r}
pvalue_over_4hr <- null_dist_over_4hrs %>% 
  get_p_value(obs_stat = obs_stat_over_4, direction = "right")


pvalue_over_4hr
```


## Wait times over 8 hours

h0: waits over 8 hours were not significantly greater in winter quarters relative to non-winter quarters
h1: the proportion of waits over 8 hours was greater in winter quarters

```{r}

waiting_times_condense %>% 
  ggplot(aes(y = number_over8hours_episode, x = winter, fill = winter)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
library(infer)

null_dist_8hrs <- waiting_times_condense %>% 
  specify(number_over8hours_episode ~ winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
```

```{r}
obs_stat_8 <- waiting_times_condense %>% 
  specify(number_over8hours_episode ~ winter) %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
            
obs_stat_8
```

```{r}
null_dist_8hrs %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_8, direction = "both")
  
```

```{r}
pvalue_8hr <- null_dist_8hrs %>% 
  get_p_value(obs_stat = obs_stat_8, direction = "right")


pvalue_8hr
```

## Wait over 12 hours

h0: waits over 12 hours were not significantly greater in winter quarters relative to non-winter quarters
h1: the proportion of waits over 12 hours was greater in winter quarters

```{r}
waiting_times_condense %>% 
  ggplot(aes(y = number_over12hours_episode, x = winter, fill = winter)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
library(infer)

null_dist_12hrs <- waiting_times_condense %>% 
  specify(number_over12hours_episode ~ winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
```

```{r}
obs_stat_over_12 <- waiting_times_condense %>% 
  specify(number_over12hours_episode ~ winter) %>% 
  calculate(stat = "diff in means", order = c("Winter", "Non-Winter"))
            
obs_stat_over_12
```

```{r}
null_dist_12hrs %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_over_12, direction = "both")
  
```

```{r}
pvalue_12hr <- null_dist_12hrs %>% 
  get_p_value(obs_stat = obs_stat_over_12, direction = "right")


pvalue_12hr
```


# H testing pre/post- covid


## Wait within 4 hours

h0: waits within 4 hours were not significantly different in post-covid quarters
h1: the proportion of waits within 4 hours was lower in post-covid quarters quarters

```{r}
waiting_times_condense %>% 
  filter(department_type == "Emergency Department") %>% 
  ggplot(aes(y = number_within4hours_episode, x = post_covid, fill = post_covid)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
library(infer)

null_dist_within_4hrs_pc <- waiting_times_condense %>% 
  filter(department_type == "Emergency Department") %>% 
  specify(number_within4hours_episode ~ post_covid) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c(T, F))
```

```{r}
obs_stat_within_4_pc <- waiting_times_condense %>% 
  specify(number_within4hours_episode ~ post_covid) %>% 
  calculate(stat = "diff in means", order = c(T, F))
            
obs_stat_within_4_pc
```

```{r}
null_dist_within_4hrs_pc %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_within_4_pc, direction = "left")
  
```

```{r}
pvalue_within_4hr_pc <- null_dist_within_4hrs_pc %>% 
  get_p_value(obs_stat = obs_stat_within_4_pc, direction = "left")


pvalue_within_4hr_pc
```





## Wait over 4 hours

h0: waits over 4 hours were not significantly greater in post-covid quarters
h1: the proportion of waits over 4 hours was greater in post-covid quarters quarters

```{r}

waiting_times_condense %>% 
  ggplot(aes(y = number_over4hours_episode, x = post_covid, fill = post_covid)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
null_dist_over_4hrs_pc <- waiting_times_condense %>% 
  specify(number_over4hours_episode ~ post_covid) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c(T, F))
```

```{r}
obs_stat_over_4_pc <- waiting_times_condense %>% 
  specify(number_over4hours_episode ~ post_covid) %>% 
  calculate(stat = "diff in means", order = c(T, F))
            
obs_stat_over_4_pc
```

```{r}
null_dist_over_4hrs_pc %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_over_4_pc, direction = "both")
  
```

```{r}
pvalue_over_4hr_pc <- null_dist_over_4hrs %>% 
  get_p_value(obs_stat = obs_stat_over_4_pc, direction = "both")


pvalue_over_4hr_pc
```

## Wait times over 8 hours

h0: waits within 8 hours were not significantly different in winter quarters relative to non-winter quarters
h1: the proportion of waits within 8 hours was greater in winter quarters

```{r}

waiting_times_condense %>% 
  ggplot(aes(y = number_over8hours_episode, x = post_covid, fill = post_covid)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
library(infer)

null_dist_8hrs_pc <- waiting_times_condense %>% 
  specify(number_over8hours_episode ~ post_covid) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c(T, F))
```

```{r}
obs_stat_8_pc <- waiting_times_condense %>% 
  specify(number_over8hours_episode ~ post_covid) %>% 
  calculate(stat = "diff in means", order = c(T, F))
            
obs_stat_8_pc
```

```{r}
null_dist_8hrs_pc %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_8_pc, direction = "both")
  
```

```{r}
pvalue_8hr_pc <- null_dist_8hrs %>% 
  get_p_value(obs_stat = obs_stat_8_pc, direction = "right")


pvalue_8hr_pc
```



h0: waits over 12 hours were not significantly different in winter quarters relative to non-winter quarters
h1: the proportion of waits over 12 hours was greater in winter quarters

```{r}
waiting_times_condense %>% 
  ggplot(aes(y = number_over12hours_episode, x = post_covid, fill = post_covid)) +
  geom_boxplot()+
  coord_flip()
```

```{r}
library(infer)

null_dist_12hrs_pc <- waiting_times_condense %>% 
  specify(number_over12hours_episode ~ post_covid) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c(T, F))
```

```{r}
obs_stat_over_12_pc <- waiting_times_condense %>% 
  specify(number_over12hours_episode ~ post_covid) %>% 
  calculate(stat = "diff in means", order = c(T, F))
            
obs_stat_over_12_pc
```

```{r}
null_dist_12hrs_pc %>% 
  visualise() +
  shade_p_value(obs_stat = obs_stat_over_12_pc, direction = "both")
  
```

```{r}
pvalue_12hr_pc <- null_dist_12hrs %>% 
  get_p_value(obs_stat = obs_stat_over_12, direction = "both")


pvalue_12hr_pc
```
